---
title: "A complete PMSLT analysis from spec to results"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{A complete PMSLT analysis from spec to results}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r library}
library(pmslttools)
```

## What this vignette covers

`pmslttools` walks a modeller from a small model specification to a full
proportional multistate lifetable (PMSLT) result. This vignette runs the whole
workflow on built-in mock data so you can see the moving parts before supplying
real inputs. The headline driver is `run_pmslt()`, which chains every layer into
one call.

The workflow has four stages:

1. **Specify and template** the model.
2. **Solve disease consistency** to get canonical disease inputs.
3. **Run the simulation** for business-as-usual (BAU) and each intervention arm.
4. **Read the results**: health outcomes, costs, and cost-effectiveness.

## 1. Specify the model and draft input templates

A `pmslt_spec()` records the intervention, the diseases it acts on, the
population structure, and the time horizon. `draft_input_templates()` turns the
spec into a folder of numbered CSV templates with a written guide.

For this vignette we use `generate_mock_pmslt_inputs()`, which drafts the
templates *and* fills them with illustrative values, so every required file is
ready to run.

```{r generate}
input_dir <- file.path(tempdir(), "pmslt_inputs_raw")
generate_mock_pmslt_inputs(output_dir = input_dir)
list.files(input_dir, pattern = "\\.csv$")
```

The numbered files separate concerns: population and all-cause rates
(`01`–`03`), disease epidemiology (`05`–`06`), intervention evidence
(`08`–`10`), and optional equity and cost inputs (`11`–`12`).

## 2. Run the whole pipeline

With filled inputs in place, `run_pmslt()` does everything: it solves disease
consistency, converts intervention evidence into disease deltas, ages the
population through the all-cause lifetable for BAU and each arm, and assembles
the reporting tables.

The default solver, `"dismod_slove"`, is deterministic and needs no optional
dependencies. (The more rigorous Bayesian solver, `"disbayes"`, is available via
`solver = "disbayes"` when the optional `disbayes` package is installed.)

```{r run, message = FALSE}
run <- run_pmslt(
  input_dir = input_dir,
  solver = "dismod_slove",
  horizon = 10
)

run
```

`run_pmslt()` returns a `pmslt_run` object. `summary()` prints the headline
results: BAU health outcomes and, for every intervention arm, the incremental
effect against BAU.

```{r summary}
summary(run)
```

## 3. Reading the results

The result object exposes each layer for further reporting.

**Health-adjusted life years.** `run$halys` holds the BAU summary and a
comparison per arm. HALYs are person-years lived minus years lived with
disability.

```{r halys}
run$halys$bau

run$halys$comparisons[["Tobacco tax"]]
```

**The lifetable bridge.** `run$lifetable` is the all-cause main lifetable object
with `bau`, the per-arm `interventions`, and ready-made `comparisons`. Because
the census templates are age-banded, `run_pmslt()` expands them to exact single
years internally before ageing the cohort.

```{r lifetable}
head(run$lifetable$bau[, c("time_step", "age", "sex", "person_years", "yld")])
```

## 4. Costs and cost-effectiveness

When a `12_costs.csv` file is present (or a cost table is passed via the `costs`
argument), `run_pmslt()` also attaches costs and computes incremental
cost-effectiveness ratios using the `intervention - BAU` convention.

```{r costs}
costs <- data.frame(
  age_start = 40, age_end = 80, age_label = "40-80",
  sex = rep(c("male", "female"), each = 2),
  stratum = "total",
  disease = c("CHD", "Stroke", "CHD", "Stroke"),
  disease_cost = 250,
  background_cost = 120,
  currency = "AUD", price_year = 2024,
  source = "vignette", notes = "",
  stringsAsFactors = FALSE
)

run_costed <- run_pmslt(
  input_dir = input_dir,
  solver = "dismod_slove",
  horizon = 10,
  costs = costs,
  overwrite = TRUE
)

run_costed$icers
```

A positive ICER with `icer_status == "positive_incremental_halys"` is the
familiar cost-per-HALY case; the status column flags zero or negative
incremental-HALY arms explicitly rather than reporting a misleading ratio.

## Stepping through the layers manually

`run_pmslt()` is a convenience wrapper. Every layer is also a public function,
so you can run them one at a time for teaching or finer control — for example
`solve_disease_consistency()`, `run_pmslt_interventions()`,
`run_pmslt_lifetable_interventions()`, `calculate_halys()`, and
`calculate_icers()`. Use `next_pmslt_step()` at any point to get guidance on the
recommended next call.

```{r next-step}
next_pmslt_step("disease_consistency")$recommended_function
```

## Where to go next

- Add probabilistic sensitivity analysis with `run_pmslt(..., psa = TRUE)`.
- Add an equity overlay with `run_pmslt(..., equity = TRUE)` and an
  `11_stratum_rate_ratios.csv` file.
- Replace the mock inputs with your own filled templates and re-run.
